---
title: "W2-Reproducible research"
author: "JBO"
date: "7/11/2020"
output: html_document
---


## Project overview  
This assignment consist of showing various key figures for a persons number of steps logged with 5 minute intervals over a 2 month period from 10/01/2012 to 11/30/2012.

## Needed packages  
The following packages are needed for the analysis to run:

```{r message=FALSE}  
library(dplyr)  
library(ggplot2)  
library(knitr)  
library(lubridate)  
```


### Loading and Preprocessing the data
Here we read in the given data and make sure the date column is treated as dates instead of characters.

```{r}
dat<-read.csv("activity.csv")
dat$date<-as.Date(dat$date)
```



### What is mean total number of steps taken per day?
To see how the persons steps a distributed on a daily basis we create a new data frame that conist of total number of steps taken each days. NA's are removed.


```{r message=FALSE}
df_steps_day<-dat %>%
  filter(!is.na(steps)) %>%
  mutate(date = floor_date(date)) %>%
  group_by(date) %>%
  summarize(total_steps = sum(steps))
```

With this new data frame we can plot the histogram of the total number of steps taken:

```{r}
hist(df_steps_day$total_steps,ylab="No of days", xlab = "No of steps", main = "Histogram of number of total steps taken per day")
```

We can further get the mean:
```{r}
mean(df_steps_day$total_steps)
```

and the median
```{r}
median(df_steps_day$total_steps)
```

## What is the average daily activity pattern?
We create a new data frame that provides the mean number of steps taken at each interval, disregarding NA's.

```{r message=FALSE}
df_interval<-dat %>%
  filter(!is.na(steps)) %>%
  group_by(interval) %>%
  summarize(mean_steps = mean(steps))
```

With this new data fram we can see how the avergae number of steps changes over the intervals.

```{r}
with(df_interval, plot(interval,mean_steps,type = "l",ylab="Mean number of steps per interval",
                       xlab = "Interval", main = "Line plot of mean number of steps taken for given interval"))
```

The interval with the highest number of mean steps can be found via:

```{r}
df_interval$interval[which.max(df_interval$mean_steps)]
```

## Imputing missing values
First we find how many NA's the dataset actually contains

```{r}
sum(is.na(dat$steps))
```


Instead of just removing NA's all together I will instead fill any missing value with the average value for the iven interval and day.  
To this end I use the weekdays function to distinguish between days and create a lookup data fram with the mean number of steps taken per interval and day.

```{r message=FALSE}
df_interpolate<-dat %>%
  filter(!is.na(steps)) %>%
  mutate(day_no = weekdays(date)) %>%
  group_by(interval,day_no) %>%
  summarize(mean_steps = mean(steps))
```

Having this lookup table available I can loop through the raw data frame and exchange the NA's with the values from the just calculated data frame.

```{r}
dat_int<-dat
for (i in 1:nrow(dat)){
  if (is.na(dat$steps[i])){
    dat_int$steps[i]<-df_interpolate$mean_steps[which(df_interpolate$day_no==weekdays(dat$date[i]) 
                                                  & df_interpolate$interval==dat$interval[i])]         
  }
}
```  

As earlier we can now find the total number of steps taken
```{r message=FALSE}
df_steps_day1<-dat_int %>%
  mutate(date = floor_date(date)) %>%
  group_by(date) %>%
  summarize(total_steps = sum(steps))
```  
and the histogram of those.

```{r}
hist(df_steps_day1$total_steps,ylab="No of days", xlab = "No of steps", main = "Histogram of number of total steps taken per day")  
```  

Mean total number of steps
```{r}
mean(df_steps_day1$total_steps)
```
and median
```{r}
median(df_steps_day1$total_steps)
```

The scheme for dealing with NA's have not changed the values completely.
## Are there differences in activity patterns between weekdays and weekends?
Last we want to see if there any differences between weekdays and weekend days.  
For that purpose we construct a data frame with a column that characterize each day as either a weekday or weekend day.
I use the wday function that can tell me by a numeric value to which category a given day belong. Mean number of steps is the calculated for each category.

As earlier we can now find the total number of steps taken
```{r message=FALSE}
df_week<-dat_int %>%
  mutate(date = floor_date(date)) %>%
  mutate(day_type = if_else(wday(date,week_start = 1)<6,"Weekday","Weekend")) %>%
  group_by(interval,day_type) %>%
  summarize(mean_steps = mean(steps))
```

A plot using the facet system from ggplot2 can now be shown.
```{r}
ggplot(df_week, aes(x=interval, y=mean_steps))+geom_line() +facet_grid(day_type~.)
```

From these graphs it seems fair to say that the persons activity level depends on it being either a weekday or weekend.


